k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-05 paper.Rmd
another round of editing

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, two simulation studies show that k-plus anticlustering is well-suited to create groups having minimal differences according to multiple distribution characteristics, significantly outperforming traditional anticlustering criteria such as the diversity and the classical k-means objective. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups instead of only means. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ that can be maximized in order to minimize differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to create groups that have similar variances across the $P$ attributes. As a caveat, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group mean $\overline{\mathbf{x}}^{(k)}$, but the additional variables computed to minimize differences with regard to the variance across groups are calculated using the overall mean $\overline{\mathbf{x}}$. Thus, $\mathit{SSE_{Var}}$ is only valid if (approximately) $\overline{\mathbf{x}}$ = $\overline{\mathbf{x}}^{(k)}$; otherwise we may introduce non-negligible measurement error when optimizing $\mathit{SSE_{Var}}$. Thus, only considering the new variables $\mathbf{Y}_{N \times P}$ in order to minimize discrepancies in variance may not yield satisfying results; instead, it has to be ensured that the group means of the original variables are also similar between groups. Fortunately, minimizing discrepancy with regard to group means can be accomplished by standard k-means anticlustering, i.e., the maximization of $\mathit{SSE}$. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

An exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is not feasible---and most likely not interesting on its own---without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). The most simple approach for an optimization that considers multiple criteria at the same time is the weighted sum approach [@naidu2014; @marler2010]. That is, all criteria are computed independently from each other and are then combined into a single criterion through a weighted sum; a suitable optimization algorithm is then used to maximize the combined criterion. By combining $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ in a single criterion, we define the objective function for k-plus anticlustering:

$$
\mathit{SSE_{kplus}} = 
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$
Maximizing $\mathit{SSE_{kplus}}$ will lead to similar means and variances between $K$ groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two-dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. A first small scale simulation compared different approaches for weight selection for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. In a second large scale simulation study, the k-plus objective was compared to other popular anticlustering objectives, i.e., the k-means and diversity objectives.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. Specifically, the initial grouping can also be chosen in such a way that the groups are different sized; for example, if the data set cannot evenly be split, it is possible to ensure that the group sizes differ by 1 at most. However, the data sets in the simulation studies were always partitioned into equal-sized groups.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constituting criteria, averaged across the `r length(results)` simulation runs. For the weighted approaches, the criterion that received the highest weight during the optimization was fulfilled to the highest degree, as should be expected. Interestingly, the unweighted standardization approach had the best average performance across the four criteria. While the degree of advantage by itself does not seem to be remarkable, this pattern was observed in an astonishing `r mean(unstd_best) * 100`% of all simulation runs, while the level of chance would have been `r 1/6 * 100`% (`r format_p(sig_test$p.value)`). Among the competing weighting schemes, the unweighted standardization approach therefore yielded the most promising results. It also yielded satisfying results with regard to maximizing the standard $\mathit{SSE}$, even slightly outperforming the weighting scheme that explicitly favored $\mathit{SSE}$. Hence, the unweighted standardization approach seemed to implicitly give high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order moment $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was used as the benchmark implementation for k-plus anticlustering that was compared to the competing objectives. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the degree to which partitions identified by the different anticlustering methods showed similarity. Specifically, I investigated whether different anticlustering solutions showed any overlap regarding the placement of individual elements in groups. In cluster analysis, high quality partitions according to different objectives tend to overlap to a large degree, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each data set and for each $K$.

@@

It is maybe surprising that generalized k-plus does not seem to have performed much better than diversity anticlustering with regard to skewness and kurtosis, even though the objective has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, generalized k-plus was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, generalized k-plus was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-rounder method that tends to address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. The Appendix includes additional figures splitting the simulation results according to the other variables that varied between simulation runs (number of features; correlation between features; standard deviation of features; number of groups).

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions according to the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitionings.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different, as was also shown in Simulation 2. In contrast, different clustering solutions tend to look similar because a good clustering is always characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This result was observed in Simulation 2 and can also be clarified using Figure 1. Here, we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.

@@

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the identification of groups with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data [@delBarrio2022], while anticlustering divides data sets into artificial subgroups in such a way that between-group similarity / within-group heterogeneity is maximized. However, these methods are logically similar because anticlustering corresponds to fair clustering when only considering the protected attributes. Hence, a fruitful outlook for research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.

>>>>

To optimize overall between-group similarity---and hence to overcome the limitations of k-means anticlustering---this paper introduces k-plus anticlustering. K-plus is an extension of the k-means objective that quantifies between-group similarity as discrepancy with regard to several distribution characteristics instead of only the mean. Specifically, k-plus offers the possibility to minimize differences with regard to the variance and higher order moments such as skewness and kurtosis. After formally introducing the k-plus objective, two simulation studies show that k-plus anticlustering is well-suited to create groups having minimal differences according to multiple distribution characteristics, significantly outperforming traditional criteria such as the diversity and the k-means objective. Practical examples illustrate how readers can easily apply k-plus anticlustering using the free and open source software package `anticlust` [@papenberg2020; @R-anticlust], an extension to the popular statistical programming language R [@R-base].

@@

In the following, I present a variation of the $\mathit{SSE}$ that can be used to minimize differences in variances between groups. To motivate the new criterion, we first consider a set of one-dimensional observations. The sample variance of a set of $N$ observations $x = (x_1, \ldots, x_N)$ is computed as the mean of the squared distances between observations and their mean $\overline{\mathbf{x}}$:

@@

Thus, to obtain an adaptation of the $\mathit{SSE}$ whose maximization can result in minimal differences in variances between groups, a new variable is computed for each original variable: The new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ with $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of $\mathbf{Y}_{N \times P}$, we define the criterion $\mathit{SSE_{Var}}$:

@@

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. However, the input data reflects the squared difference between each data point and the attribute mean instead of the raw data points. Maximizing $\mathit{SSE_{Var}}$ can be used to obtain groups that have similar variances across the $P$ attributes. As a caveat, however, it is crucial to note that maximizing $\mathit{SSE_{Var}}$ only reliably yields similar group variances if the group centroids $\overline{\mathbf{x}}^{(k)}$ ($k = 1, \dots, K$) are close to the overall data centroid $\overline{\mathbf{x}}$. This is because the within-group variance is calculated on the basis of the group centroid $\overline{\mathbf{x}}^{(k)}$, but the new squared distance variables $y_{ij}$ are calculated using the overall centroid $\overline{\mathbf{x}}$ as reference point.  To ensure that $\mathit{SSE_{Var}}$ is a valid criterion for between-group similarity in variance, we must therefore simultaneously ensure that group centroids of the original variables are close to the overall data centroid (and therefore to each other). Ensuring that the group means are close to the overall centroid is accomplished when maximizing the classical $\mathit{SSE}$, i.e., via standard k-means anticlustering [@spath1986]. Combining the maximization of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ at the same time leads to the k-plus approach for anticlustering, which is detailed in the next section.

@@

An exclusive maximization of $\mathit{SSE_{Var}}$ (i.e., equalizing variances) is not feasible---and most likely not interesting on its own---without also considering the standard $\mathit{SSE}$ (i.e., equalizing means). The most simple approach for an optimization that considers multiple criteria at the same time is the weighted sum approach [@naidu2014; @marler2010]. That is, all criteria are computed independently from each other and are then combined into a single criterion through a weighted sum. By combining $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ in a single criterion, we define the objective function for k-plus anticlustering:

$$
\mathit{SSE_{kplus}} = 
w_1 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{x}_j ^{(k)})^2 + 
w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

Maximizing $\mathit{SSE_{kplus}}$ will lead to similar means and variances between $K$ groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($t = 1, \dots, T$), leading to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two-dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. A first small scale simulation compared different approaches for weight selection for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. In a second large scale simulation study, the k-plus objective was compared to other popular anticlustering objectives, specifically, the diversity and the k-means objectives.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping directly determines the resulting group sizes, because the following exchange procedure never changes the sizes of the groups. Hence, while the data sets in the simulation studies were always partitioned into equal-sized groups, it is possible to specify different sized groups through the initial assignment, or if the data set cannot evenly be split, it is possible to ensure that the group sizes differ by 1 at most.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The results of this initial test determined the benchmark implementation of k-plus anticlustering in the follow-up large scale comparison in Simulation 2.

In Simulation 1, the target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Six weighting schemes were compared. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardised data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

Table 1 illustrates how well each of the six weighting schemes fulfilled each of the four constituting criteria, averaged across the `r length(results)` simulation runs. For the weighted approaches, the criterion that had received the highest weight during the optimization was fulfilled to the highest degree, as should be expected. Interestingly, the unweighted standardization approach had the best average performance across the four criteria. While the degree of advantage by itself does not seem to be remarkable, this pattern was observed in an astonishing `r mean(unstd_best) * 100`% of all simulation runs, while the level of chance would have been `r 1/6 * 100`% (`r format_p(sig_test$p.value)`). Among the competing weighting schemes, the unweighted standardization approach therefore yielded the most promising results. It also yielded satisfying results with regard to maximizing the standard $\mathit{SSE}$, even slightly outperforming the weighting scheme that explicitly favoured $\mathit{SSE}$. Hence, the unweighted standardization approach implicitly gave high priority to minimizing differences in means. This is a pleasant result because, as discussed earlier, for the higher order moment $\mathit{SSE}$s to be valid, the group means have to be close to the overall mean, which is best accomplished if the standard $\mathit{SSE}$ has high priority.

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach that performed best in Simulation 1 was employed. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the standard k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

@@

As a secondary purpose of Simulation 2, I investigated the similarity of partitions obtained when optimizing the different anticlustering objectives. In cluster analysis, partitions identified through different approaches tend to overlap considerably, but it is an open question whether this result also holds for different anticlustering partitions. To tackle this issue, I computed the adjusted Rand index (ARI) for each pair of partitions that were returned by the competing methods [@rand1971; @hubert1985]. The ARI is an index of the similarity between two partitions, where an index of 0 indicates no similarity and a value of 1 indicates the highest possible similarity (i.e., when two partitions are the same). The pairwise ARIs were computed for each simulation run, i.e., for data set and for each $K$.

@@

It is maybe surprising that generalized k-plus does not seem to have performed much better than diversity anticlustering with regard to skewness and kurtosis, even though the objective has been specifically tailored to these criteria. Figure 2 gives a more fine grained assessment of this observation: Surprisingly, for small $N$, generalized k-plus was even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperformed diversity anticlustering. These results illustrate the cost of optimizing several criteria at the same time: For low $N$, generalized k-plus was apparently preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitated to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. Generally, it should be noted that diversity anticlustering is a good all-rounder method that tends to address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering was outperformed. The Appendix includes additional figures splitting the simulation results according to the other variables that varied between simulation runs (number of features; correlation between features; standard deviation of features; number of groups).

@@

Table 3 illustrates the degree of overlap between anticlustering partitions that were identified in Simulation 2. Overall, it is shown that there was hardly any similarity between anticlustering partitions that were obtained via optimization of the different criteria. The average ARI was almost 0 in all cases---indicating no similarity of partitions whatsoever. Moreover, the ARI between two anticlustering partitions hardly differed from the ARI between an anticlustering partition and a random partition. This constitutes an interesting result, illustrating that high between-group similarity can be achieved via very different partitions.

@@

Anticlustering is usually implemented by reversing an objective function used in classical cluster analysis. Whereas cluster analysis seeks homogeneity within groups and separation between groups, the reversal results in heterogeneity within groups and similarity between groups. While tightly linked in a mathematical sense, an important semantic difference is that cluster analysis is concerned with finding "true" structure in the data, e.g., by defining a taxonomy on naturally occurring groups [@everitt1979]. Anticlustering instead imposes structure on the data, while the anticlusters themselves do not represent any underlying "true" pattern. Correspondingly, different---but similarly adequate---anticlustering partitions tend to look very different. In contrast, different clustering solutions tend to look similar because good clusterings are characterized by assigning similar elements into the same cluster and dissimilar elements into different clusters. A good anticlustering solution however does not depend on assigning individual items either to the same or a different group, independent of their pairwise dissimilarity.[^dispersion] This result was observed in Simulation 2 and can also be clarified using Figure 1. Here, we observe a high concentration of suitable anticlustering partitions but only few suitable clustering solutions. An astonishing `r mean(df$kmeans / max(df$kmeans) >.999) * 100`% of all partitions are within 0.1% of the global maximum in $\mathit{SSE}$, i.e., the best k-means anticlustering solution. In contrast, there are only `r sum(min(df$kmeans)/ df$kmeans > .90)` partitions (`r mean(min(df$kmeans)/ df$kmeans > .90)*100`%) that approximate the global minimum in $\mathit{SSE}$ (i.e., the best k-means clustering solution) by even 10%. Thus, even though the problems are based on the same objective function, cluster analysis and anticlustering exhibit a stark discrepancy with regard to the solution space of partitions. Interestingly, the k-plus objective introduces an additional distinction between cluster analysis and anticlustering. Reversing the k-plus anticlustering objective no longer yields a useful clustering interpretation, because identifying groups that are different with regard to higher order moments does not constitute a clustering. Thus, even though k-plus extends the classical k-means clustering criterion, it is a pure anticlustering objective.

@@

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the identification of groups with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data, while anticlustering divides data sets into artificial subgroups in such a way that between-group similarity / within-group heterogeneity is maximized. However, these methods are logically similar because anticlustering corresponds to fair clustering when only considering the protected attributes. Hence, a fruitful outlook for research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.
